---
title: "Reproducible Research: Peer Assessment 1"
author: "Pinecode99"
date: "October 23, 2016"
output: html_document
keep_md: true
---

#Loading and preprocessing the data
Assuming you have manually downloaded the data, you can load it with
the following R code:
```{r}
library(dplyr)
library(lubridate)
act1 <- read.csv("./activity.csv", header = TRUE, sep = ",")

```


#What is mean total number of steps taken per day?
Part 1 answer. Total steps taken per day is calculated using:
```{r, echo=TRUE}
act2 <- aggregate(steps~date,act1,sum)
```


Part 2 Answer
Use 15 bins (breaks) to display adequately:
```{r, echo=TRUE}
hist(act2$steps,breaks=15, main="Histogram of mean steps/day", xlab="Mean steps/day")
```


Part 3 answer
Create a mean file and median file then combine them
```{r, echo=TRUE}
act3 <- as.data.frame(mean(act2$steps))
names(act3) <- "mean"
act4 <- as.data.frame(median(act2$steps))
names(act4) <- "median"
act5 <- cbind(act3, act4)
act5
# rm(act2,act3,act4)
```


#What is the average daily activity pattern?
Make a time series plot (i.e. type = "l" ) of the 5-minute interval (x-axis) and the average
```{r, echo=TRUE}
act10 <- aggregate(steps~interval,act1,mean)
plot(steps~interval,data=act10, type="l", xlab="Interval", main="Average number of steps taken by interval (Oct-Nov 2012)")
```


Which 5-minute interval, on average across all the days in the dataset,  
contains the maximum number of steps?
```{r, echo=TRUE}
MaxInt1 <- which.max(act10$steps)
MaxInt2 <- act10[MaxInt1,]
MaxInt2
```


#Imputing missing values
Calculate and report the total number of missing values in the dataset (i.e. the total
number of rows with NA's)
```{r, echo=TRUE}
sum(is.na(act1[,1])) #2304
sum(is.na(act1[,2])) #0
sum(is.na(act1[,3])) #0
```

Devise a strategy for filling in all of the missing values in the dataset. The strategy
does not need to be sophisticated. My strategy will be to use the mean of the relevant segments.


Create a new dataset that is equal to the original dataset but with the missing data filled in.
create a new copy from act1 to work with
```{r, echo=TRUE}
act30 <- act1
rm(act1)

for (i in 1:nrow(act30))
{
       if (is.na(act30$steps[i]))
       {
         #build code to lookup mean data based on the interval
         interval1 <- act30[i,3]
         interval2 <- which(act10$interval==interval1)
         meansteps <- act10[interval2,2]
         act30[i,1] <- meansteps
       }
        
}

rm(act10)
# sum(is.na(act30[,1])) # demosstrates the NA's have been replaced

```


Make a histogram of the total number of steps taken each day and Calculate and report the
mean and median total number of steps taken per day. Do these values differ from the estimates 
from the first part of the assignment? What is the impact of imputing missing data on the estimates
of the total daily number of steps?
```{r, echo=TRUE}
act31 <- aggregate(steps~date,act30,sum)
hist(act31$steps,breaks=15, main="Histogram of mean steps/day", xlab="Mean steps/day")
```

create a file for mean and one for median and combine them
```{r, echo=TRUE}
act32 <- as.data.frame(mean(act31$steps))
names(act32) <- "mean"
act33 <- as.data.frame(median(act31$steps))
names(act33) <- "median"
act34 <- cbind(act32, act33)
```

this is the result (coincidentally! (or not) the mean and median are forced to be
the same using the method choosen) this can be demonstrated by sorting the act31 column by steps
```{r, echo=TRUE}
act34
rm(act31,act32,act33,act34)
```


answer: the results differ by a small amout. Using averaged data for the missing data 'tightened up' the histogram
making the data more centrally located than would have been the case if we had the real data.



#Are there differences in activity patterns between weekdays and weekends?
For this part the weekdays()  function may be of some help here. Use the dataset with the
filled in missing values for this part

Create a new factor variable in the dataset with two levels weekday and weekend
indicating whether a given date is a weekday or weekend day.
 
convert the date filed from factor to date
```{r, echo=TRUE}
act40 <- act30
rm(act30)
act40$date <- as.Date(act40$date)
act41 <- mutate(act40, daycode = weekdays(date))
act42 <- mutate(act41,code="a")

for (i in 1:nrow(act42))
{
        if (act42$daycode[i] == "Saturday" | act42$daycode[i] == "Sunday")
        {
               act42$code[i] <- "weekend"
        } 
        else
        {
                act42$code[i] <- "weekday"
        }
        
}
head(act42,5)
rm(act41)
```


```{r, echo=TRUE}
par(mfrow=c(2,1))

### a plot for weekdays
act50 <- filter(act42,code=="weekday")
act51 <- aggregate(steps~interval,act50,mean)
plot(steps~interval,data=act51, type="l", xlab="Interval", main="Average number of steps (weekday)")

### a plot for weekends
act55 <- filter(act42,code=="weekend")
act56 <- aggregate(steps~interval,act55,mean)
plot(steps~interval,data=act56, type="l", xlab="Interval", main="Average number of steps (weekend)")

###dev.off()

rm(act40,act42,act50,act51,act55,act56)

```

